HL-LHC Computing Review:\\Common Tools and Community Software

### 2020-04-30 LHCC/2020/2020-01/ctcs.tex
Upload LHCC Common Tools and Community Spftware document (#131)
Upload "final" version into GitHub

N.B. This version is squashed - editing history is in lhcc-document branch in my fork

%% Comment out before final submission %%
\usepackage{lineno}  % for line numbering during review
\linenumbers

@@

% HSF Document Number (only for arXiv)
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-DOC-2020-01 \\
 & & Zenodo DOI Here \\
 & & May 1, 2020 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

>>>>

%% Comment out before final submission %%
\usepackage{lineno}  % for line numbering during review
%\linenumbers

@@

% HSF Document Number (only for arXiv)
\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-DOC-2020-01 \\
 & & 10.5281/zenodo.3779250 \\
 & & May 1, 2020 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

### 2020-04-30 LHCC/2020/2020-01/input_recoTrigger.tex
Upload LHCC Common Tools and Community Spftware document (#131)
Upload "final" version into GitHub

N.B. This version is squashed - editing history is in lhcc-document branch in my fork

At the HL-LHC, the central challenge for high-level triggers (e.g.,
software triggers) and object reconstruction is to maintain excellent
efficiency and resolution in the face of high pileup values, especially
at low transverse momenta. Detector upgrades such as increases in
channel density, high precision timing and improved detector layouts are
essential to overcome these problems. The subsequent increase of event
complexity at the HL-LHC also requires the development of software
algorithms that can process events with a similar cost per event to
Run-2 and Run-3. At the same time, algorithmic approaches need to
effectively take advantage of evolutions in computing technologies,
including increased SIMD capabilities, increased core counts in CPUs,
and heterogeneous hardware.

This section focuses on the challenges identified in the Community White
Paper~\cite{Alves:2017she,albrecht2018hepexec,albrecht2018hep} and the development of solutions that. It also includes
mentions of open source software efforts that have developed within or
outside LHC collaborations that could be adapted by the LHC community to
improve trigger and reconstruction algorithms.

@@

Trigger systems are evolving to be more capable, both in their ability
to select a wider range of events of interest for the physics program of
their experiment, and their ability to stream a larger rate of events
for further processing. The event rates that will be processed by
experiments at the HL-LHC will increase by up to a factor 10 with
respect to Run-3, owing to upgraded trigger systems and expanded physics
programs. ATLAS and CMS plan to maintain a two-tiered trigger system,
where a hardware trigger makes use of coarse event information to reduce
the event rate to 10x over the current capability, up to 1 MHz \cite{ATLAS-TDR-29,collaboration:2714892}. The high level trigger system, implemented in software, selects
up to 100 kHz of events to be saved in full for subsequent
analysis\footnote{LHCb \cite{Aaij:2019uij} and ALICE \cite{Buncic:2011297} will both
  stream the full collision rate to real-time or quasi-realtime software
  trigger systems.}.

Experiments have also been working towards minimising the differences
between trigger (online) and offline software for reconstruction and
calibration, so that more refined physics objects can be obtained
directly within the HLT farm for a more efficient event selection. This
is also in-line with enhancing the experiment's capabilities for
real-time data analysis of events accepted by the hardware trigger
system, driven by use cases where the physics potential improves when
analysing more events than what can be written out in full with
traditional data processing.

Implementations of real-time analysis systems, where raw data is
processed into its final form as close as possible to the detector (e.g.
in the high-level trigger farm), are in use within several experiments.
These approaches remove the detector data that typically makes up the
raw data tier kept for offline reconstruction, and keep only a limited
set of analysis objects reconstructed within the high level trigger.

Real-time analysis techniques are being adopted to enable a wider range
of physics signals to be saved by the trigger for final analysis. As
rates increase, these techniques can become more important and
widespread by enabling only the parts of an event associated with the
signal candidates to be saved, reducing the required disk space. The
experiments are focusing on the technological developments that make it
possible to do this with acceptable reduction of the analysis
sensitivity and with minimal biases. The HSF trigger and reconstruction
group has been encouraging cross-talk between LHC experiments and beyond
on real-time analysis, as these kinds of non-standard trigger and
reconstruction workflows increase the physics output for selected
physics cases without adding significant overhead to the current
resources.

Active research topics include compression and custom data formats;
toolkits for real-time detector calibration and validation which will
enable full offline analysis chains to be ported into real-time; and
frameworks which will enable non-expert offline analysts to design and
deploy real-time analyses without compromising data taking quality.
Use-cases for real-time analysis techniques have expanded during the
last years of Run 2 and their use appears likely to
grow already during Run 3 data taking for all experiments owing to
improvements in the HLT software and farms \textbf{Cite CMS and ATLAS CDRs}. Further ideas include retaining very reduced data from as much
of the data taking rate as possible from selected regions and
sub-detectors, even up to the 40 MHz bunch crossing rate~\cite{hannes_sakulin_2019_3598769}.

@@

Algorithmic areas of particular importance at HL-LHC experiments are
charged-particle trajectory reconstruction (\textit{tracking}), including
hardware triggers based on tracking information which may seed later
software trigger and reconstruction algorithms; jet reconstruction,
including the use of high-granularity calorimetry; and the use of
precision timing detectors.

@@

The CPU needed for event reconstruction in Run-2 and Run-3 has been
dominated by charged particle reconstruction (tracking). This is still a
focus for HL-LHC triggering and event reconstruction, especially when
the need for efficiently reconstructing low transverse momentum
particles is considered.

The huge increase in the number of charged particles and hence the
combinatorial load for tracking algorithms at future colliders will put
great strain on compute resources. To minimise the memory footprint,
tracking software needs to be thread-safe to support multi-threaded
execution per core. At the same time, the software has to be efficient
and accurate to meet the physics requirements.

Since the CWP, experiments have made progress towards improving software
tracking efficiency in HL-LHC simulation, see e.g. Ref. \cite{ATL-PHYS-PUB-2019-041}. A
number of collaboration and community initiatives have been focusing on
tracking, targeting both offline and online, and we list them below.

The \emph{\textbf{ACTS}} project \cite{Ai:2019kze,Gumpert_2017} is an attempt to encapsulate
the current ATLAS track reconstruction software into an
experiment-independent and framework-independent tracking software
designed to fully exploit modern computing architectures. It builds on
the tracking experience already obtained at the LHC, and targets the
HL-LHC as well as future hadron colliders. ACTS provides a set of track
reconstruction tools designed for parallel architectures, with a
particular emphasis on thread-safety and concurrent event
reconstruction. ACTS has active collaborators from other HEP
experiments, such as FASER, and provides support to Belle-II, EIC, CEPC,
etc.

The aim of the \textbf{mkFit} project is to speed up Kalman filter (KF)
tracking algorithms using highly parallel architectures and to deliver
track building (and possibly fitting) software for the HL-LHC. Recent
activities focused on delivering a production quality software setup to
perform track building in the context of the LHC Run-3 (or Run-2). The
implementation relies on single-precision floating point mathematics and is
available for multicore CPUs and achieves significant gains in compute
performance from the use of vector instructions set extending to
AVX-512, based on Matriplex library (a part of the mkFIt project).

@@

Physics performance in very high pileup environments, such as the HL-LHC
or FCC-hh, may also benefit from adding timing information to the
reconstruction. This allows the mitigation of the effects of pile-up by
exploiting the time-separation of collision products~\cite{Collaboration:2623663,CMS:2667167}\\
The experiment communities have been working on timing detector
reconstruction and object identification techniques in complex
environments. Since the CWP, initial tracking and vertexing algorithms
that include timing information have been developed and incorporated
into experimental software stacks \cite{LindseyGreyMIPCTD}\footnote{Tracking algorithms where events overlap in time
within time slices are employed by the CBM experiment~\cite{Akishina:2015ghv,AkishinaThesis}. The use of a cellular automaton makes the
  algorithm less dependent on the specific detector geometry, and there
  may be the possibility of cross-talk with LHC experiments that are
  investigating 4D reconstruction.}.

@@

HEP experiments have extensive continuous integration systems, including
code regression checks that have enhanced the monitoring procedures for
software development in recent years. They also have automated
procedures to check trigger rates as well as the performance of the low-
and high-level physics objects in the data {[}ref ATLAS DQ and CMS
DQ{]}.

@@

Improving the ability of HEP developed toolkits to use vector units on
commodity hardware will bring speedups to applications running on both
current and future hardware. The goal for work in this area is to evolve
current toolkit and algorithm implementations, and best programming
techniques to better use SIMD capabilities of current and future
computing architectures. Since the CWP, algorithm development projects
(e.g. \cite{LHCB-FIGURE-2019-002}
as well as mkFit) have demonstrated success in increasing the use of vector
units. In addition, HEP has increasingly benefited from industry
developed software, including machine learning toolkits (e.g.,
TensorFlow), that typically make excellent use of vector units.
Challenges remain in this area. For example, full exploitation of this
must also account for many generations of hardware that experiments must
exploit on the grid. In addition, to realise full performance gains, a
large fraction of the application must be improved to use SIMD processor
capabilities (partly due to turbo-boost capabilities of processors)

@@

It may be desirable or even necessary to deploy new algorithms,
including advanced machine learning techniques, in order to face the
increase in event complexity without increasing the per-event
reconstruction resource needs.

Work is already undergoing in the collaborations to evolve or rewrite
existing toolkits and algorithms focused on their physics and technical
performance at high event complexity (e.g. high pileup at HL-LHC), and
efforts in this area have expanded. Cross-collaboration developments are
focusing on the availability of ML algorithms for experimental software,
especially in persistifying models and running inferences in production
and in enhancing the
training capacities of collaborations by submitting jobs to the grid and
to facilities offering large CPU/GPU resources~\cite{LWTNN, ONNX}. One can also foresee
that HEP will increase the benefit from techniques and cross-talk from
industry.

@@

Computing platforms are generally evolving towards having more cores in
order to increase processing capability. In addition different
architectural models, e.g. GPUs, FPGAs are available.
The goal is to improve the throughput of multithreaded
software trigger and event reconstruction applications, so current
event models, toolkits and algorithm implementations have to evolve to
efficiently exploit these opportunities. The first algorithms that should be
targetted are those which are most time consuming.

Since the CWP, HEP now has a number of algorithmic projects that are
designed for, or have successfully adapted to, hardware accelerators,
most notably to NVidia GPUs using CUDA. Two projects targeting Run-3 and
with prospects for use in Run-4 are Patatrack (from CMS) \cite{andrea_bocci_2019_3598824} and
Allen (from LHCb) \cite{Aaij:2019zbu}.

\textbf{Patatrack} is a CMS initiative aimed to use heterogeneous
computing for the charged particle reconstruction. Within the CMS
codebase (CMSSW), this project has demonstrated that physics
reconstruction code can be written to leverage heterogeneous
architectures, like GPUs, to achieve a significant speed-up while
reducing the overall cost and power consumption of a system. An example
is the CMS Pixel local reconstruction and the track and vertex
reconstruction: these algorithms can run on an NVIDIA T4 GPU with the
same performance as 52-56 Xeon cores, at a fraction of the cost and
power consumption of an equivalent CPU-only system; further improvements
may come pairing one or more high end GPUs with a low power ARM system.

\textbf{Allen} is a data processing framework for GPUs, as well as a
specific implementation of a first-level trigger for LHCb Run 3
data-taking. Allen is optimised for sustaining the rate required by
real-time processing in LHCb, but can be used as a more general GPU data
processing as it includes a scheduler and a memory manager that can be
integrated into Gaudi. This permits the extension and implementation of
Allen within other experimental software using the same underlying
framework.

ATLAS is also investigating GPU solutions for reconstruction software
and its acceleration at HL-LHC, in particular in terms of highly
parallel algorithms \cite{attila_krasznahorkay_2019_3599103} and to enable the use of machine learning
algorithms in object reconstruction and identification {[}\emph{waiting
for an email from ATLAS with references}{]}.

The experience acquired with these projects shows that GPUs can be used
for increasing the throughput in cases of large-size (ALICE) \cite{david_rohr_2019_3599418} and small-size events per second (LHCb), and to speed up
specific parts of the data processing (CMS, Patatrack).

@@

Heterogeneous computing is still a rapidly developing and changing
field. The burden of writing architecture-specific code repeatedly will
certainly exceed the available effort from the experiments. Efforts are
ongoing in this direction, since the CWP, trying to take advantage of
the latest development in synergy with industry and attempting to
increase the use of centres with heterogeneous resources by experiments.

@@

The use of open-source elements of the LHC software stack for trigger
and reconstruction by smaller experiments, especially in case of common
experimental design\footnote{An example of software sharing happens in
  the case of the FASER experiment, whose offline software
  (\href{https://gitlab.cern.ch/faser/calypso}{{https://gitlab.cern.ch/faser/calypso}})
  is based on a derivative of the ATLAS Athena open-source software, which in turn uses the Gaudi framework. FASER
  also shares tracking detector components with ATLAS, facilitating the
  adoption of software relying on similar detector descriptions.}, is
still worthwhile This benefits experiments with a smaller developer base
than LHC experiments, and increases the return on investment for LHC
experiment software. For this reason, we also encourage the addition of
common open software projects to the \href{https://projectescape.eu/services/open-source-scientific-software-and-service-repository}{ESCAPE software catalogue}.

>>>>

At the HL-LHC, the central challenge for high-level triggers (e.g.
software triggers) and object reconstruction is to maintain excellent
efficiency and resolution in the face of high pileup values, especially
at low transverse momenta. Detector upgrades, such as increases in
channel density, high precision timing and improved detector layouts are
essential to overcome these problems. The subsequent increase of event
complexity at the HL-LHC also requires the development of software
algorithms that can process events with a similar cost per-event to
Run-2 and Run-3. At the same time, algorithmic approaches need to
effectively take advantage of evolutions in computing technologies,
including increased SIMD capabilities, increased core counts in CPUs,
and heterogeneous hardware.

This section focuses on the challenges identified in the Community White
Paper~\cite{Alves:2017she,albrecht2018hepexec,albrecht2018hep} and the
development of solutions since then. It also includes
mentions of open source software efforts that have developed within or
outside LHC collaborations that could be adapted by the LHC community to
improve trigger and reconstruction algorithms.

@@

Trigger systems are evolving to be more capable, both in their ability
to select a wider range of events of interest for the physics program of
their experiment, and their ability to stream a larger rate of events
for further processing. The event rates that will be processed by
experiments at the HL-LHC will increase by up to a factor 10 with
respect to Run-3, owing to upgraded trigger systems and expanded physics
programs. ATLAS and CMS plan to maintain a two-tiered trigger system,
where a hardware trigger makes use of coarse event information to reduce
the event rate to 10x over the current capability, up to 1 MHz \cite{ATLAS-TDR-29,collaboration:2714892}. The high level trigger system, implemented in software, selects
up to 100 kHz of events to be saved in full for subsequent
analysis\footnote{LHCb \cite{Aaij:2019uij} and ALICE \cite{Buncic:2011297} will both
  stream the full collision rate to real-time or quasi-real-time software
  trigger systems.}.

Experiments have also been working towards minimising the differences
between trigger (online) and offline software for reconstruction and
calibration, so that more refined physics objects can be obtained
directly within the HLT farm for a more efficient event selection. This
is also in-line with enhancing the experiment's capabilities for
real-time data analysis of events accepted by the hardware trigger
system, driven by use cases where the physics potential improves when
analysing more events than can be written out in full with
traditional data processing.

Implementations of real-time analysis systems, where raw data is
processed into its final form as close as possible to the detector (e.g.
in the high-level trigger farm), are in use within several experiments.
These approaches remove the detector data that typically makes up the
raw data kept for offline reconstruction, and keep only a limited
set of analysis objects reconstructed within the high level trigger or
keepig only the parts of an event associated with the
signal candidates, reducing the required disk space.
The experiments are focusing on the technological developments that make it
possible to do this with acceptable reduction of the analysis
sensitivity and with minimal biases. The HSF Reconstruction and Software Trigger
group has been encouraging cross-talk between LHC experiments and beyond
on real-time analysis, as these kinds of workflows increase the physics output for selected
physics cases without adding significant overhead to the current
resources.

Active research topics also include compression and custom data formats;
toolkits for real-time detector calibration and validation which will
enable full offline analysis chains to be ported into real-time; and
frameworks which will enable non-expert offline analysts to design and
deploy real-time analyses without compromising data taking quality.
Use cases for real-time analysis techniques have expanded during the
last years of Run 2 and their use appears likely to
grow already during Run 3 data taking for all experiments owing to
improvements in the HLT software and farms. % Missing reference to CMS and ATLAS CDRs
Further ideas include retaining high-rate, but very reduced data, from selected regions and
sub-detectors, even up to the 40 MHz bunch crossing rate~\cite{hannes_sakulin_2019_3598769}.

@@

Algorithmic areas of particular importance for HL-LHC experiments are
charged-particle trajectory reconstruction (\textit{tracking}), including
hardware triggers based on tracking information which may seed later
software trigger and reconstruction algorithms; jet reconstruction,
including the use of high-granularity calorimetry; and the use of
precision timing detectors.

@@

The CPU needed for event reconstruction in Runs 2 and 3 is
dominated by charged particle reconstruction (tracking). This is still a
focus for HL-LHC triggering and event reconstruction, especially when
the need to efficiently reconstruct low transverse momentum
particles is considered.

The huge increase in the number of charged particles, and hence the
combinatorial load for tracking algorithms, at future colliders will put
great strain on compute resources. To minimise the memory footprint,
tracking software needs to be thread-safe and to support multi-threaded
execution per core. At the same time, the software has to be efficient
and accurate to meet the physics requirements.

Since the CWP, experiments have made progress towards improving software
tracking efficiency in HL-LHC simulation, e.g. \cite{ATL-PHYS-PUB-2019-041}. A
number of collaboration and community initiatives have been focusing on
tracking, targeting both offline and online, and these are listed below.

The \textbf{ACTS} project \cite{Ai:2019kze,Gumpert_2017} is an attempt to encapsulate
the current ATLAS track reconstruction software into an
experiment-independent and framework-independent tracking software,
designed to fully exploit modern computing architectures. It builds on
the tracking experience already obtained at the LHC, and targets the
HL-LHC as well as future hadron colliders. ACTS provides a set of track
reconstruction tools designed for parallel architectures, with a
particular emphasis on thread-safety and concurrent event
reconstruction. ACTS has active collaborators from other HEP
experiments, such as FASER, and provides support to Belle-II, EIC and CEPC.

The aim of the \textbf{mkFit} \cite{cerati2019speeding} project is to speed up Kalman filter (KF)
tracking algorithms using highly parallel architectures and to deliver
track building (and possibly fitting) software for the HL-LHC. Recent
activities focused on delivering a production quality software setup to
perform track building in the context of LHC Runs 2 and 3. The
implementation relies on single-precision floating point mathematics and is
available for multicore CPUs. It achieves significant gains in compute
performance from the use of vector instructions, extending to
AVX-512, based on Matriplex library (a part of the mkFit project).

@@

Physics performance in very high pileup environments, such as the HL-LHC
or FCC-hh, may also benefit from adding timing information to the
reconstruction. This allows the mitigation of the effects of pile-up by
exploiting the time-separation of collision products~\cite{Collaboration:2623663,CMS:2667167}

Experiment communities have been working on timing detector
reconstruction and object identification techniques in complex
environments. Since the CWP, initial tracking and vertexing algorithms
that include timing information have been developed and incorporated
into experimental software stacks \cite{LindseyGreyMIPCTD}\footnote{Tracking algorithms where events overlap in time
within time slices are employed by the CBM experiment~\cite{Akishina:2015ghv,AkishinaThesis}. The use of a cellular automaton makes the
  algorithm less dependent on the specific detector geometry, and there
  may be the possibility of cross-talk with LHC experiments that are
  investigating 4D reconstruction.}.

@@

HEP experiments have extensive continuous integration systems, including
code regression checks that have enhanced the monitoring procedures for
software development in recent years. They also have automated
procedures to check trigger rates as well as the performance of the low-
and high-level physics objects in the data. % Missing reference {[}ref ATLAS DQ and CMS DQ{]}.

@@

Improving the ability of HEP developed toolkits to use vector units on
commodity hardware will bring speedups to applications running on both
current and future hardware. The goal for work in this area is to evolve
current toolkit and algorithm implementations, and develop best programming
techniques to better use the SIMD capabilities of current and future
computing architectures. Since the CWP, algorithm development projects
have demonstrated success in increasing the use of vector
units~\cite{cerati2019speeding, LHCB-FIGURE-2019-002}. In addition, HEP has increasingly benefited from industry
developed software, including machine learning toolkits (e.g.,
TensorFlow), that typically make excellent use of vector units.
Challenges remain in this area: for example, full exploitation
must also account for many generations of hardware that experiments must
exploit on the grid. In addition, to realise full performance gains, a
large fraction of the application must be improved to use SIMD processing
capabilities (partly due to turbo-boost capabilities of processors).

@@

It may be desirable, or even necessary, to deploy new algorithms that
include advanced machine learning techniques to manage the
increase in event complexity without increasing per-event
reconstruction resource needs.

Work is already ongoing in the collaborations to evolve or rewrite
existing toolkits and algorithms focused on their physics and technical
performance at high event complexity (e.g. high pileup at HL-LHC), and
efforts in this area have expanded. Cross-collaboration developments are
focusing on the availability of ML algorithms for experimental software,
especially in persistifying models and running inferences in production
and in enhancing the
training capacities of collaborations by submitting jobs to the grid and
to facilities offering large CPU/GPU resources~\cite{LWTNN, ONNX}. One can also foresee
that HEP will increase the benefit from techniques and cross-talk from
industry.

@@

Computing platforms are generally evolving towards having more cores
or to adopt different
architectural models (GPUs, FPGAs) to increase processing capability~\cite{acm10.1145/3282307}.
The goal for HEP is to improve the throughput of
software trigger and event reconstruction applications, so current
event models, toolkits and algorithm implementations have to evolve to
efficiently exploit these opportunities. The first algorithms that should be
targeted are those which are most time consuming.

Since the CWP, HEP now has a number of algorithmic projects that are
designed for, or have successfully adapted to, hardware accelerators,
most notably to NVIDIA GPUs using CUDA. Two projects targeting Run-3 and
with prospects for use in Run-4 are Patatrack (from CMS) \cite{andrea_bocci_2019_3598824} and
Allen (from LHCb) \cite{Aaij:2019zbu}.

\textbf{Patatrack} is a CMS initiative aimed at using heterogeneous
computing for charged particle reconstruction. Within the CMS
codebase (CMSSW), this project has demonstrated that physics
reconstruction code can be written to leverage heterogeneous
architectures, like GPUs, and achieve a significant speed-up, while
reducing the overall cost and power consumption of a system. An example
is the CMS Pixel local reconstruction and the track and vertex
reconstruction: these algorithms can run on an NVIDIA T4 GPU with the
same performance as 52-56 Xeon cores, at a fraction of the cost and
power consumption of an equivalent CPU-only system; further improvements
may come pairing one or more high end GPUs with a low power ARM system.

\textbf{Allen} is a data processing framework for GPUs, as well as a
specific implementation of a first-level trigger for LHCb Run-3
data-taking. Allen is optimised for sustaining the rate required by
real-time processing in LHCb, but can be used as a more general GPU data
processing system as it includes a scheduler and a memory manager that can be
integrated into Gaudi~\cite{gaudi_paper}. This permits the extension and implementation of
Allen within other experimental software using the same underlying
framework.

ATLAS is also investigating GPU solutions for reconstruction software
and its acceleration at HL-LHC, in particular in terms of highly
parallel algorithms \cite{attila_krasznahorkay_2019_3599103} and to enable the use of machine learning
algorithms in object reconstruction and identification. % Missing references {[}\emph{waiting for an email from ATLAS with references}{]}.

The experience acquired with these projects shows that GPUs can be used
for increasing the throughput in cases of large-size (ALICE)
\cite{david_rohr_2019_3599418} and small-size events per second (LHCb), and to speed up
specific parts of the data processing (CMS Patatrack).

@@

%Heterogeneous computing is still a rapidly developing and changing
%field. The burden of writing architecture-specific code repeatedly will
%certainly exceed the available effort from the experiments. Efforts are
%ongoing in this direction, since the CWP, trying to take advantage of
%the latest development in synergy with industry and attempting to
%increase the use of centres with heterogeneous resources by experiments.

@@

The use of open-source elements of the LHC software stacks for trigger
and reconstruction by smaller experiments, especially in case of common
experimental design\footnote{An example of software sharing happens in
  the case of the FASER experiment, whose offline software
  (\href{https://gitlab.cern.ch/faser/calypso}{{https://gitlab.cern.ch/faser/calypso}})
  is based on a derivative of the ATLAS Athena open-source software, which in turn uses the Gaudi framework. FASER
  also shares tracking detector components with ATLAS, facilitating the
  adoption of software relying on similar detector descriptions.}, is
still worthwhile This benefits smaller experiments
and increases the return on investment for LHC
experiment software. For this reason, we also encourage the addition of
common open software projects to the \href{https://projectescape.eu/services/open-source-scientific-software-and-service-repository}{ESCAPE software catalogue}.

### 2020-04-30 LHCC/2020/2020-01/input_summary.tex
Upload LHCC Common Tools and Community Spftware document (#131)
Upload "final" version into GitHub

N.B. This version is squashed - editing history is in lhcc-document branch in my fork

The main challenges and R\&D lines of interest have been discussed
above. There are clearly identified priority topics for the different
areas in the next few years. We note that longer term R\&D, into areas
such as quantum or neuromorphic computing, should happen at a low level,
but are almost certainly too far from production to deliver improvements
for Run 4 of the LHC.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Undertake a detailed performance analysis of current production
  detector simulation code, to better understand where performance
  limitations in data and cache locality arise from.
  \end{quote}
\item
  \begin{quote}
  Improve current codes through refactoring and specialised HEP models
  to avoid bottlenecks identified,
  \end{quote}
\item
  \begin{quote}
  Develop further fast simulation models and techniques, including
  machine learning based options in addition to optimised parametric
  approaches as well as common frameworks for their tuning and
  validation.
  \end{quote}
\item
  \begin{quote}
  Develop GPU codes for tackling very specific, time consuming, parts of
  the simulation for HEP. Specifically calorimetry, including geometry,
  field and electro-magnetic physics processes.
  \end{quote}
\item
  \begin{quote}
  Develop integration prototypes to exercise and benchmark the
  simultaneous use of GPU specific libraries with CPU-based software
  (e.g. Geant4) to cover all particles and processes required in
  experiments simulation frameworks.
  \end{quote}
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Develop fast and performant reconstruction software, making optimal
  use of hardware resources. A particularly crucial development for
  HL-LHC is that of pattern recognition solutions (including timing
  information, if available) that can withstand the increase of pile-up
  at HL-LHC. Wherever possible, solutions should be uniform for online
  and offline software, so the experiments can benefit from consistent
  event selection between trigger and final analysis and from more
  effective real-time analysis results.
  \end{quote}
\item
  \begin{quote}
  Develop solutions that can exploit accelerators (FPGA, GPU) for
  trigger and reconstruction, including general-purpose interoperability
  libraries with a focus to generalising and mitigating their
  dependencies on architecture-specific codes.
  \end{quote}
\item
  \begin{quote}
  Further promote activities in the area of data quality and software
  monitoring, as the quality of the recorded/reconstructed data is
  paramount for all physics analysis.
  \end{quote}
\item
  \begin{quote}
  Support efforts towards the implementation of machine learning models
  in experimental framework (C++) to enable more widespread use in
  trigger and reconstruction software.
  \end{quote}
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Identify a CPU-efficient data reduction and storage model that serves
  the majority of analyses with a footprint of O(1kb/event), retaining
  the flexibility to accommodate the needs of specialised analyses.
  \end{quote}
\item
  \begin{quote}
  Streamline analysis metadata access and calibration schemes, and
  provide effective book-keeping for fractional datasets.
  \end{quote}
\item
  \begin{quote}
  Develop cross-experimental tools establishing declarative syntax
  and/or languages for analysis description, interfaced with distributed
  computing backends.
  \end{quote}
\item
  \begin{quote}
  Define and improve schemes for interoperability of end-stage
  experimental software with data science/machine learning frameworks.
  \end{quote}
\end{enumerate}

Undertaking this programme of priority topics and R\&D requires, first
and foremost, investment from funding agencies to support developments.
A new generation of developers is also needed to refresh and regenerate
the long lived software projects on which the field so heavily relies.
The transition from R\&D to production ready software is usually long
and so it is urgent to undertake this investment now, with the positive
outcomes ready to become integrated in time for HL-LHC. Regenerating the
cadre of software developers in HEP also requires support for realistic
long term career prospects for those who specialise in this vital area.
Collaboration with industry can certainly be fruitful and is to be
encouraged to allow early access to technology and communication of HEP
problems and priorities.

@@

The stakeholders in HEP must be able to give their priorities and
feedback on R\&D areas. Each project has mechanisms for doing this,
while the HSF can help to provide overall input on prioritisation.

>>>>

The main challenges and R\&D lines of interest have been discussed
above. There are clearly identified priority topics for the different
areas in the next few years\footnote{Longer term R\&D, into high-risk, high-reward blue sky areas,
such as quantum or neuromorphic computing, should happen at a low level,
but are almost certainly too far from production to deliver improvements
for Run 4 of the LHC.}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Undertake a detailed performance analysis of current production
  detector simulation code, to better understand where performance
  limitations in data and cache locality arise from.
  \end{quote}
\item
  \begin{quote}
  Improve current codes through refactoring and specialised HEP models
  to avoid bottlenecks identified.
  \end{quote}
\item
  \begin{quote}
  Develop further fast simulation models and techniques, including
  machine learning based options in addition to optimised parametric
  approaches as well as common frameworks for their tuning and
  validation.
  \end{quote}
\item
  \begin{quote}
  Undertake R\&D into GPU codes for tackling very specific, time consuming, parts of
  the simulation for HEP. Specifically calorimetry, including geometry,
  field and electro-magnetic physics processes.
  \end{quote}
\item
  \begin{quote}
  Develop integration prototypes to exercise and benchmark the
  simultaneous use of GPU specific libraries with CPU-based software
  (e.g. Geant4) to cover all particles and processes required in
  experiments simulation frameworks.
  \end{quote}
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Develop fast and performant reconstruction software, making optimal
  use of hardware resources. A particularly crucial development for
  HL-LHC is that of pattern recognition solutions (including timing
  information, if available) that can withstand the increase of pile-up
  at HL-LHC. Wherever possible, solutions should be uniform for online
  and offline software, so the experiments can benefit from consistent
  event selection between trigger and final analysis and from more
  effective real-time analysis results.
  \end{quote}
\item
  \begin{quote}
  Develop solutions that can exploit accelerators (FPGA, GPU) for
  trigger and reconstruction, including general-purpose interoperability
  libraries with a focus to generalising and mitigating their
  dependencies on architect\-ure specific codes.
  \end{quote}
\item
  \begin{quote}
  Further promote activities in the area of data quality and software
  monitoring, as the quality of the recorded and reconstructed data is
  paramount for all physics analysis.
  \end{quote}
\item
  \begin{quote}
  Support efforts towards the implementation of machine learning models
  in experimental framework (C++) to enable more widespread use in
  trigger and reconstruction software.
  \end{quote}
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Identify a CPU-efficient data reduction and storage model that serves
  the majority of analyses with a footprint of O(1kb/event), retaining
  the flexibility to accommodate the needs of specialised analyses.
  \end{quote}
\item
  \begin{quote}
  Streamline analysis metadata access and calibration schemes, and
  provide effective book-keeping for fractional datasets.
  \end{quote}
\item
  \begin{quote}
  Develop cross-experimental tools establishing declarative syntax
  and/or languages for analysis description, interfaced with distributed
  computing backends.
  \end{quote}
\item
  \begin{quote}
  Define and improve schemes for interoperability of end-stage
  experimental software with data science and machine learning frameworks.
  \end{quote}
\end{enumerate}

Undertaking this programme of priority topics and R\&D requires, first
and foremost, investment from funding agencies to support developments.
A new generation of developers is also needed to refresh and regenerate
the long lived software projects on which the field so heavily relies.
The transition from R\&D to production ready software is usually long
and so it is urgent to undertake this investment now, with the positive
outcomes ready to become integrated in time for HL-LHC. Regenerating the
cadre of software developers in HEP also requires support for realistic
long term career prospects for those who specialise in this vital area.
Collaboration with industry can certainly be fruitful and is to be
encouraged to allow early access to technology and communication of HEP
problems and priorities.
R\&D that is successful will become the next generation of production
software, but this will require lifetime support for maintainance and
further evolution.

@@

The stakeholders in HEP must be able to give their priorities and
feedback on R\&D areas. Each project has mechanisms for doing this,
while the HSF can help to provide overall input on prioritisation
and coordination of common cross-experiment activities.
